---
editor_options:
  markdown:
    wrap: 80
---

# `main_fn`

The fn below is under no title or subtitle in the pkgdown index.

```{r}
main_fn <- function(...) {...}
```

# First group of fns

## DESCRIPTION

The description text for *First group of fns*.

## `group_wide_fn`

The fn below is only under a title, but no subtitle in the pkgdown index.

```{r}
group_wide_fn <- function(...) {...}
```

## Some subgroup

The fns below are both under a title and subtitle in the pkgdown index.

### `subgroup_fn_1`

```{r}
subgroup_fn_1 <- function(...) {...}
```

### `subgroup_fn_2`

```{r}
subgroup_fn_2 <- function(...) {...}
```

### Sub-subgroup

The above *Sub-subgroup* heading is ignored for the pkgdown index because the
latter only supports up to two heading levels.

The fn below thus moves one hierarchy level up in the pkgdown index.

#### `sub_subgroup_fn`

```{r}
sub_subgroup_fn <- function(...) {...}
```

## Another subgroup

### DESCRIPTION

The description text for *Another subgroup*.

### `another_subgroup_fn`

```{r}
another_subgroup_fn <- function(...) {...}
```

# Second group of fns

## Yet another subgroup

### `yet_another_subgroup_fn`

```{r}
yet_another_subgroup_fn <- function(...) {...}
```
