---
editor_options:
  chunk_output_type: console
---

# NOTES

-   This file is *not* included in the [source package](https://r-pkgs.org/structure.html#sec-source-package) because of the [`.nopurl` suffix in its
    filename](https://pkgpurl.rpkg.dev/reference/purl_rmd.html#-rmd-files-excluded-from-purling).

-   The chunks below have to be manually executed in order to regenerate the package data.

-   Although the datasets below are saved as ["internal data" in `R/sysdata.rda`](https://r-pkgs.org/data.html#sec-data-sysdata), they can still be exported and
    documented (by documenting the dataset's quoted name in the main `.Rmd` source file -- which only works when the dataset is also `@export`ed), something
    [not explicitly mentioned](https://coolbutuseless.github.io/2018/12/10/r-packages-internal-and-external-data/) in the book [R
    Packages](https://r-pkgs.org/data.html#sec-data-data). To do so, you first need to manually add the `export()` directive in the NAMESPACE file since
    roxygen2 won't add it automatically.

# Define data

## `pkg_config`

```{r}
pkg_config <-
  tibble::tibble(key = character(),
                 default_value = list(),
                 description = character()) |>
  tibble::add_row(key = "add_copyright_notice",
                  description = paste0("Whether or not to add a **copyright notice** at the beginning of the generated `.R` files as recommended by e.g. the ",
                                       "[GNU licenses](https://www.gnu.org/licenses/gpl-howto.html). The notice consists of the name and description of the ",
                                       "program and the word `Copyright (C)` followed by the release years and the name(s) of the copyright holder(s), or if ",
                                       "not specified, the author(s). The year is always the current year. All the other information is extracted from the ",
                                       "package's `DESCRIPTION` file.")) |>
  tibble::add_row(key = "add_license_notice",
                  description = paste0("Whether or not to add a **license notice** at the beginning of the generated `.R` files as recommended by e.g. the ",
                                       "[GNU licenses](https://www.gnu.org/licenses/gpl-howto.html). The license is determined from the package's ",
                                       "`DESCRIPTION` file and currently only the [`AGPL-3.0-or-later` ",
                                       "license](https://spdx.org/licenses/AGPL-3.0-or-later.html) is supported.")) |>
  tibble::add_row(key = "gen_pkgdown_ref",
                  description = paste0("Whether or not to overwrite [pkgdown](https://pkgdown.r-lib.org/)'s [reference ",
                                       "index](https://pkgdown.r-lib.org/reference/build_reference.html#reference-index) in the configuration file ",
                                       "`_pkgdown.yml` with an auto-generated one based on the main input file as described in [pkgpurl::gen_pkgdown_ref()]."))
```

# Write data

Save all the small data objects as a single internal file `R/sysdata.rda`. Note that when documenting them, they must be explicitly `@export`ed to be available
to package users.

```{r}
usethis::use_data(pkg_config,
                  internal = TRUE,
                  overwrite = TRUE,
                  compress = "xz",
                  version = 3L)
```
