---
output: pal::gitlab_document
---

# `r paste0(pal::desc_value("Package"), ": ", pal::desc_value("Title"))`

`r pal::desc_value("Description")`

The basic idea behind the concept this package implements originates from Yihui Xie. See his blog post
[_Write An R Package Using Literate Programming Techniques_](https://yihui.org/rlp/) for more details, it's definitively worth reading. This package's function
`pkgpurl::purl_rmd()` is just a less cumbersome alternative to the Makefile approach outlined by him.

## Details

Besides the ability to write accompanying information in expressive [Markdown](https://en.wikipedia.org/wiki/Markdown), the
[R Markdown](https://rmarkdown.rstudio.com/) format provides several further advantages over the bare R source format:

- If you use RStudio or any other editor with proper R Markdown syntax highlighting, you will probably like the gained visual clarity for distinguishing
  individual functions/code parts (i.e. R code chunks). This also facilitates creating a meaningful document structure (in Markdown) alongside the actual R
  source code.

- You can use RStudio's [code outline](https://rviews.rstudio.com/2016/11/11/easy-tricks-you-mightve-missed/#code-outline) to easily navigate through longer
  scripts. It provides significantly better usability than the
  [code section standard](https://support.rstudio.com/hc/en-us/articles/200484568-Code-Folding-and-Sections) of classic R scripts. It makes it easy to find
  your way around source files that are thousands of lines long.

- You can have development-only code which never lands in the generated R source files (and thus the R package) in separate code chunks with the chunk option
  [`purl = FALSE`](https://yihui.org/knitr/options/#extracting-source-code). This turns out to be very convenient in certain situations.

- As you might already know, [you can only use ASCII characters in your code if you plan on submitting your package to CRAN](https://r-pkgs.org/r.html#r-cran).
  This of course also applies to R code chunks in the `Rmd/*.Rmd` files that are compiled to `R/*-GEN.R` files by pkgpurl. But since the accompanying Markdown
  documentation won't land in the compiled R files, you can use the full Unicode spectrum of characters there (🥳) – as long as you exclude the R Markdown
  source from the built package by putting a line `^Rmd$` in `.Rbuildignore` (which is recommended anyway).

But there are also a few drawbacks of the R Markdown format:

- The pkgpurl approach on writing R packages in the R Markdown format introduces _one_ additional step at the very beginning of typical package development
  workflows: Running `pkgpurl::purl_rmd()` to generate the `R/*-GEN.R` files from the original `Rmd/*.Rmd` sources before documenting/checking/testing/building
  the package. Given sufficient user demand, this could probably be integrated into [devtools](https://devtools.r-lib.org/)' functions in the future, so that
  no additional action has to be taken by the user when relying on RStudio's built-in package building infrastructure.

  For the time being, it's recommended to set up a custom shortcut[^recommended-keycombo] for `pkgpurl::purl_rmd()` which is registered as an
  [RStudio add-in](https://rstudio.github.io/rstudioaddins/).

- Setting up a new project to write an R package in the R Markdown differs slightly from the classic approach. A suitable convenience function like
  `create_rmd_package()` to set up all the necessary parts could probably be added to [usethis](https://usethis.r-lib.org/) in the future.

  For the time being, you can use my ready-to-go [R Markdown Package Development Template](https://gitlab.com/salim_b/r/pkg-dev-tpl) as a starting point for
  creating new R packages in the R Markdown format.

- Other than in `.R` files, RStudio currently doesn't support auto-completion of [roxygen2 tags](https://roxygen2.r-lib.org/articles/rd.html) in `.Rmd` files.
  The same applies to the roxygen2 comment continuation when inserting a newline. These are [known issues](https://github.com/rstudio/rstudio/issues/5809)
  which will hopefully be resolved in the near future.


[^recommended-keycombo]: I personally recommend to use <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd> since it's not occupied yet by any of the predefined
[RStudio shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts).


## Installation
```{r, install-snippet-dev, child = pkgsnippets::rmd_snippet_path("installation-notice_dev-version_gitlab.Rmd")}
```

## Development

### R Markdown format
```{r, rmd-pkg-snippet, child = pkgsnippets::rmd_snippet_path("rmd-package-notice.Rmd")}
```

### Coding style
```{r, coding-style-snippet, child = pkgsnippets::rmd_snippet_path("coding-style-notice.Rmd")}
```

---
