---
output: pal::gitlab_document
---

```{r}
#| label: init
#| include: false

knitr::opts_knit$set(root.dir = getwd())
```

# `r pal::desc_value("Package")`

`r paste0('[![CRAN Status](https://r-pkg.org/badges/version/', pal::desc_value(key = "Package"), ')](https://cran.r-project.org/package=', pal::desc_value(key = "Package"), '){.pkgdown-release}')`

`r pal::desc_value("Description")`

The basic idea behind the concept this package implements originates from Yihui Xie. See his blog post [*Write An R Package Using Literate Programming
Techniques*](https://yihui.org/rlp/) for more details, it's definitively worth reading. This package's function
[`pkgpurl::purl_rmd()`](https://pkgpurl.rpkg.dev/dev/reference/purl_rmd.html) is just a less cumbersome alternative to the Makefile approach outlined by him.

## Pros and cons

The [R Markdown](https://rmarkdown.rstudio.com/) format provides several **advantages** over the bare R source format when developing an R package:

-   👍 **Mix Markdown and Code**

    It allows the actual code to be freely mixed with explanatory and supplementary information in expressive [Markdown](https://en.wikipedia.org/wiki/Markdown)
    format instead of having to rely on [`#` comments](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Comments) only. In general, this should
    encourage to actually record code-accompanying information because you're able to use the full spectrum of [Pandoc's Markdown
    syntax](https://pandoc.org/MANUAL.html#pandocs-markdown) like inline formatting, lists, tables, quotations or math[^1].

    It is especially powerful in combination with the [Visual R Markdown](https://rstudio.github.io/visual-markdown-editing/) feature introduced in RStudio 1.4,
    which -- in addition to the visual editor -- offers a feature whose utility can hardly be overestimated: Pandoc Markdown
    [canonicalization](https://en.wikipedia.org/wiki/Canonicalization) (on file save[^2]). For example, it allows paragraphs being wrapped automatically at the
    desired line width; or to write a minimal sloppy [pipe table](https://pandoc.org/MANUAL.html#extension-pipe_tables) that is automatically normalized to a
    beautifully formatted and actually readable one.

    The relevant editor options which adjust the canonical Markdown generation can either be set

    -   [per `.Rmd` file](https://rstudio.github.io/visual-markdown-editing/#/markdown?id=writer-options), e.g.

        ``` rmd
        ---
        editor_options:
          markdown:
            wrap: 160
            references:
              location: section
            canonical: true
        ---
        ```

    -   or [per project](https://rstudio.github.io/visual-markdown-editing/#/options?id=project-options) in the usual `PACKAGE_NAME.Rproj` file, e.g.

        ``` ini
        MarkdownWrap: Column
        MarkdownWrapAtColumn: 160
        MarkdownReferences: Section
        MarkdownCanonical: Yes
        ```

        (I'd recommend to set them *per project*, so they apply to the whole package including any `.Rmd` vignettes.)

-   👍 **All your code in a single, well-structured file**

    The [traditional recommendation](https://style.tidyverse.org/package-files.html) to not lose overview of your package's R source code is to split it over
    multiple files. The popular (and very useful) book *R Packages* [gives the following advice](https://r-pkgs.org/code.html#sec-code-organising):

    > If it's very hard to predict which file a function lives in, that suggests it's time to separate your functions into more files or reconsider how you are
    > naming your functions and/or files.

    I think this is just ridiculous.

    Instead, I encourage you to keep all your code (as far as possible) in a single file `Rmd/PACKAGE_NAME.Rmd` and structure it according to the [rules
    described here](https://pkgpurl.rpkg.dev/dev/reference/gen_pkgdown_ref.html#details), which even allows the [pkgdown `Reference:`
    index](https://pkgdown.r-lib.org/reference/build_reference.html#reference-index) to be automatically in sync with the source code structure. As a result,
    you re-organize (and thus most likely improve) your package's code structure whenever you intend to improve the pkgdown reference -- and vice versa. For a
    basic example, see [this very package's main source file](https://gitlab.com/rpkg.dev/pkgpurl/-/blob/master/Rmd/pkgpurl.Rmd?plain=0).

    Keeping all code in a single file frees you from the traditional hassle of finding a viable (but in the end still unsatisfactory) way to organize your R
    source code across multiple files. Of course, there are still good reasons to outsource code into separate files *in certain situations*, which nothing is
    stopping you from doing. You can also [exclude whole `.Rmd` files from purling using the `.nopurl.Rmd` filename
    suffix](https://pkgpurl.rpkg.dev/dev/reference/purl_rmd.html#-rmd-files-excluded-from-purling).

-   👍 **Improved overview and navigation**

    You can rely on RStudio's [code outline](https://rviews.rstudio.com/2016/11/11/easy-tricks-you-mightve-missed/#code-outline) to easily navigate through
    longer `.Rmd` files. IMHO it provides significantly better usability than the [code section
    standard](https://support.posit.co/hc/en-us/articles/200484568-Code-Folding-and-Sections) of `.R` files. It makes it easy to find your way around source
    files that are thousands of lines long.

    RStudio's [*Go to File/Function* shortcut](https://support.posit.co/hc/en-us/articles/200711853-Keyboard-Shortcuts) works the same for `.Rmd` files as it
    does for `.R` files.

-   👍 **Improved visual clarity**

    If you use RStudio or any other editor with proper R Markdown syntax highlighting, you will probably like the gained visual clarity for distinguishing
    individual functions/code parts (by putting them in separate R code chunks). This also facilitates creating a meaningful document structure (in Markdown)
    alongside the actual R source code.

-   👍 **Easily toggle code inclusion**

    You can put development-only code which never lands in the generated R source files (and thus the R package) in separate code chunks with the chunk option
    [`purl = FALSE`](https://yihui.org/knitr/options/#extracting-source-code). This turns out to be very convenient in certain situations.

    For example, this is a good way to reproducibly document the generation of cleaned versions of [exported data](https://r-pkgs.org/data.html#sec-data-data)
    as well as [internal data](https://r-pkgs.org/data.html#sec-data-sysdata). This avoids having to outsource the code to separate files under `data-raw/` and
    adding the directory to `.Rbuildignore`, i.e. no need to use `usethis::use_data_raw()`. Instead, you just set `purl = FALSE` for the relevant code chunk(s).
    You can (and should) still use [`usethis::use_data()`](https://usethis.r-lib.org/reference/use_data.html) (optionally with `overwrite = TRUE`) to generate
    the files under `data/` holding external package data as well as the `R/sysdata.rda` file (using `internal = TRUE`) holding internal package data.

-   👍 **Easily toggle styler**

    If you use [styler](https://styler.r-lib.org/) to auto-format your code globally by [setting
    `knitr::opts_chunk$set(tidy = "styler")`](https://styler.r-lib.org/articles/third-party-integrations.html), you can still opt-out on a per-chunk basis by
    setting [`tidy = FALSE`](https://github.com/r-lib/styler/releases/tag/v1.5.1). This gives pleasant flexibility.

Unfortunately, there are also a few notable **drawbacks** of the R Markdown format:

-   👎 **Additional workflow step**

    The pkgpurl approach on writing R packages in the R Markdown format introduces *one* additional step at the very beginning of typical package development
    workflows: Running [`pkgpurl::purl_rmd()`](https://pkgpurl.rpkg.dev/dev/reference/purl_rmd.html) to generate the `R/*.gen.R` files from the original
    `Rmd/*.Rmd` sources before documenting/checking/testing/building the package. Given sufficient user demand, this could probably be integrated into
    [devtools](https://devtools.r-lib.org/)' functions in the future, so that no additional action has to be taken by the user when relying on RStudio's
    built-in package building infrastructure.

    For the time being, it's recommended to set up a custom shortcut[^3] for one or both of
    [`pkgpurl::purl_rmd()`](https://pkgpurl.rpkg.dev/dev/reference/purl_rmd.html) and
    [`pkgpurl::process_pkg()`](https://pkgpurl.rpkg.dev/dev/reference/process_pkg.html) which are registered as [RStudio
    add-ins](https://rstudio.github.io/rstudioaddins/).

-   👎 **Differing setup**

    Setting up a new project to write an R package in the R Markdown differs slightly from the classic approach. A suitable convenience function like
    `create_rmd_package()` to set up all the necessary parts could probably be added to [usethis](https://usethis.r-lib.org/) in the future.

    For the time being, you can use my ready-to-go [R Markdown Package Development Template](https://gitlab.com/salim_b/r/pkg-dev-tpl) as a starting point for
    creating new R packages in the R Markdown format.

-   👎 **Unwieldy debugging**

    Debugging can be a bit more laborious since line numbers in warning and error messages always refer to the generated `R/*.gen.R` file(s), not the underlying
    `Rmd/*.Rmd` source code file(s). If need be, you first have to look up the line numbers in the `R/*.gen.R` file(s) to understand which function / code parts
    cause the issue in order to know where to fix it in the `Rmd/*.Rmd` source(s).

-   👎 **Missing roxygen2 auto-completion**

    Other than in `.R` files, RStudio currently doesn't support auto-completion of [roxygen2 tags](https://roxygen2.r-lib.org/articles/rd.html) in `.Rmd` files
    and its <kbd>Reflow Comment</kbd> command doesn't properly work on them. These are [known
    issues](https://github.com/rstudio/rstudio/issues/5809#issuecomment-932228146) which will hopefully be resolved in the near future.

```{r}
#| label: documentation
#| eval: !expr '!isTRUE(getOption("pal.build_readme.is_pkgdown"))'
#| results: asis
#| echo: false
pkgsnip::md_snip(id = "pkgdown_notice") %>%
  paste0("## Documentation\n\n",
         "[![Netlify Status](https://api.netlify.com/api/v1/badges/f639cb7f-f5fe-454c-b20d-1e06916daffc/deploy-status)]",
         "(https://app.netlify.com/sites/pkgpurl-rpkg-dev/deploys)\n\n",
         .) %>%
  pal::cat_lines()
```

[^1]: Actually, you could write anything you like in any syntax outside of R code chunks as long as you don't mind the file to be *knittable* (which it doesn't
    have to be).

[^2]: It basically sends the (R) Markdown file on a "Pandoc round trip" on every file save.

[^3]: I personally recommend to use the shortcut <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd> since it's not occupied yet by any of the predefined [RStudio
    shortcuts](https://support.posit.co/hc/en-us/articles/200711853-Keyboard-Shortcuts).

## Installation

```{r}
#| label: install-snippet-dev
#| child: !expr pkgsnip::snip_path("installation-notice_dev-version_gitlab.Rmd")
```

```{r}
#| label: usage
#| eval: !expr isTRUE(getOption("pal.build_readme.is_pkgdown"))
#| results: asis
#| echo: false
pkgsnip::md_snip(id = "usage_notice") %>%
  paste0("## Usage\n\n", .) %>%
  pal::cat_lines()
```

## Package configuration

```{r}
#| label: pkg-config
#| echo: false
#| results: "asis"

pkgsnip::md_snip(id = "pkg_config",
                 pkg = "pkgpurl")
cat("\n\n::: {.table-wide}\n ")
pal::print_pkg_config("pkgpurl",
                      roxy_to_md = TRUE)
cat("\n :::")
```

## Development

### R Markdown format

```{r}
#| label: rmd-format
#| child: !expr pkgsnip::snip_path("rmd-package-notice.Rmd")
```

### Coding style

```{r}
#| label: coding-style
#| child: !expr pkgsnip::snip_path("coding-style-notice.Rmd")
```
