---
editor_options:
  chunk_output_type: console
---

# NOTES

-   This file is *not* included in the [source package](https://r-pkgs.org/structure.html#sec-source-package) because of the [`.nopurl` suffix in its
    filename](https://pkgpurl.rpkg.dev/reference/purl_rmd.html#-rmd-files-excluded-from-purling).

-   The chunks below have to be manually executed in order to regenerate the package data.

-   When used internally, other than internal data, [exported data](https://r-pkgs.org/data.html#sec-data-data) must always be referred to by its explicit
    namespace.

# Define data

## `SOME_DATASET`

```{r}
SOME_DATASET <- SOME_R_CODE()
```

# Write data

Save all the bigger data objects to separate files under `data/*.rda`. Note that when documenting them, they mustn't be explicitly `@export`ed since they're
already automatically exported and thus available to package users.

```{r}
usethis::use_data(SOME_DATASET,
                  internal = FALSE,
                  overwrite = TRUE,
                  compress = "xz",
                  version = 3L)
```
